Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-03-13 appendix-regression.Rmd
Minor edits

Given data $(y, x)$, we _fit_ the model by finding estimates for the parameters $\alpha$, $\beta$, and $\sigma^2$ that best reflect the data under the assumption that the model is correct and according to our chosen criterion that defines 'best'.
Here we consider the least-squares estimates for the coefficients $\alpha$ and $\beta$, which are also the maximum likelihood estimates if we further assume that the errors $\varepsilon$ are normally distributed.
<!-- Defining the residuals as $\hat \varepsilon(a) = y(a) - (\hat \alpha + \hat \beta x(a))$,  -->
The least-squares estimates of $\alpha$ and $\beta$ are the values that minimize the residual sum of squares, $\text{RSS} = \sum_{a} \hat \varepsilon(a)$.
These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of $x$ and $y$,
\begin{align}
  (\#eq:lm-hat)
  \hat \alpha &= \bar y - \hat \beta \bar x \\
  \hat \beta &= \frac{s(y,x)}{s^2(x)} = r(y,x) \cdot \frac{s(y)}{s(x)}.
\end{align}
<!-- (@wasserman2004allo Chapter 13), -->
The estimate for the slope coefficient, $\hat \beta$, equals the ratio of the sample covariance of $y$ with $x$ relative to the variance in $x$ or, equivalently, the correlation of $y$ with $x$ multiplied by the ratio of their standard deviations.
The estimate for the intercept, $\hat \alpha$, is the value such that the sample means follow the regression relationship, $\bar y = \hat \alpha + \hat \beta \bar x$.
<!-- We are most interested in the estimate of the slope coefficient, $\hat \beta$ -->
The standard unbiased and maximum likelihood estimates for the variance $\sigma^{2}$ are both approximately given by the sample variance of the residuals,
\begin{align}
  (\#eq:lm-var)
  \hat{\sigma}^2(\hat \beta) \approx s^2(\hat \varepsilon).
\end{align}

@@

To understand how the measurement error in $y$ impacts the researcher's regression analysis, consider the three linear regression equations for $y$, $z$, and $d$,
\begin{align}
  (\#eq:lm-y)
  y(a) &= \alpha_y + \beta_y x(a) + \varepsilon_y(a),
\end{align}
\begin{align}
  (\#eq:lm-z)
  z(a) &= \alpha_z + \beta_z x(a) + \varepsilon_z(a),
\end{align}
\begin{align}
  (\#eq:lm-d)
  d(a) &= \alpha_d + \beta_d x(a) + \varepsilon_d(a).
\end{align}
The researcher would like to fit the model for $y$ (Equation \@ref(eq:lm-y)), but instead can only fit the model for $z$ (Equation \@ref(eq:lm-z)).
The two are related via the Equation \@ref(eq:lm-z) for $d$.
It is helpful to imagine, for the moment, that $y$ and $d$ are known and so we can fit all three models \@ref(eq:lm-y), \@ref(eq:lm-z), and \@ref(eq:lm-d).
From Equation \@ref(eq:lm-hat) and the linearity of sample means and covariances it follows that 
\begin{align}
  (\#eq:lm-hat-rel)
  \hat \alpha_z &= \hat \alpha_y + \hat \alpha_d \\
  \hat \beta_z &= \hat \beta_y + \hat \beta_d.
\end{align}
That is, the estimated coefficients for $y$, $z$, and $d$ mirror the relationship between the variables themselves,  $z = y + d$.
Consequently, we see that the measurement error $d$ creates absolute errors in the regression coefficients $\hat \alpha$ and $\hat \beta$ equal to
\begin{align}
  (\#eq:lm-hat-err)
  \text{abs. error}(\hat \alpha) &\equiv \hat \alpha_z - \hat \alpha_y = \hat \alpha_d \\
  \text{abs. error}(\hat \beta) &\equiv \hat \beta_z - \hat \beta_y = \hat \beta_d.
\end{align}
These absolute errors correspond to the _statistical bias_ in the coefficient estimates.
Expressed in terms of sample means and covariances, these errors are
\begin{align}
  (\#eq:lm-hat-err-1)
  \text{abs. error}(\hat \alpha) &= \bar d - \hat \beta_d \bar x \\
  \text{abs. error}(\hat \beta)  &= \frac{s(d,x)}{s^2(x)} = r(d,x) \cdot \frac{s(d)}{s(x)}.
\end{align}

@@

Although the absolute error is the same for each species, its practical significance varies.
Recall from Equation \@ref(eq:lm-hat) that the correct value for $\hat \beta$ (i.e., the estimate without measurement error) is
\begin{align}
  \hat \beta &= \frac{s[\log \text{abun}, x]}{s^2(x)}.
\end{align}
For species whose log abundance covaries with $x$ much more than the log mean efficiency, the error will be relatively small.
This situation can occur either because the mean efficiency varies relatively little across samples or because its variation is relatively uncorrelated with $x$.
For species whose log abundance covaries with $x$ similar or less than the log mean efficiency, the error will be significant.
A sign error occurs when the species covariance is of _equal sign and smaller in magnitude_ than the covariance of the log mean efficiency.

@@

This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate LFCs in proportions from MGS count data either with or without bias correction.

>>>>

Given data $(y, x)$, we _fit_ the model by finding estimates for the parameters $\alpha$, $\beta$, and $\sigma^2$ that best reflect the data under the assumption that the model is correct and according to our chosen criterion that defines 'best'.
Here we consider the least-squares estimates for the coefficients $\alpha$ and $\beta$, which are also the maximum likelihood estimates if we further assume that the errors $\varepsilon$ are normally distributed.
<!-- Defining the residuals as $\hat \varepsilon(a) = y(a) - (\hat \alpha + \hat \beta x(a))$,  -->
The least-squares estimates of $\alpha$ and $\beta$ are the values that minimize the residual sum of squares, $\text{RSS} = \sum_{a} \hat \varepsilon(a)^2$.
These estimates can be conveniently expressed in terms of the sample means and covariances or correlations of $x$ and $y$,
\begin{align}
  (\#eq:lm-hat)
  \hat \alpha &= \bar y - \hat \beta \bar x \\
  \hat \beta &= \frac{s(y,x)}{s^2(x)} = r(y,x) \cdot \frac{s(y)}{s(x)}.
\end{align}
<!-- (@wasserman2004allo Chapter 13), -->
The estimate for the slope coefficient, $\hat \beta$, equals the ratio of the sample covariance of $y$ with $x$ relative to the variance in $x$ or, equivalently, the correlation of $y$ with $x$ multiplied by the ratio of their standard deviations.
The estimate for the intercept, $\hat \alpha$, is the value such that the sample means follow the regression relationship, $\bar y = \hat \alpha + \hat \beta \bar x$.
<!-- We are most interested in the estimate of the slope coefficient, $\hat \beta$ -->
The standard unbiased and maximum likelihood estimates for the variance $\sigma^{2}$ are both approximately given by the sample variance of the residuals,
\begin{align}
  (\#eq:lm-var)
  \hat{\sigma}^2(\hat \beta) \approx s^2(\hat \varepsilon).
\end{align}

@@

To understand how the measurement error in $y$ impacts the researcher's regression analysis, consider the three linear regression equations for $y$, $z$, and $d$,
\begin{align}
  (\#eq:lm-y)
  y(a) &= \alpha_y + \beta_y x(a) + \varepsilon_y(a),
\end{align}
\begin{align}
  (\#eq:lm-z)
  z(a) &= \alpha_z + \beta_z x(a) + \varepsilon_z(a),
\end{align}
\begin{align}
  (\#eq:lm-d)
  d(a) &= \alpha_d + \beta_d x(a) + \varepsilon_d(a).
\end{align}
The researcher would like to fit the model for $y$ (Equation \@ref(eq:lm-y)), but instead can only fit the model for $z$ (Equation \@ref(eq:lm-z)).
The two are related via the Equation \@ref(eq:lm-z) for $d$.
It is helpful to imagine that $y$ and $d$ are known and so we can fit all three models \@ref(eq:lm-y), \@ref(eq:lm-z), and \@ref(eq:lm-d).
From Equation \@ref(eq:lm-hat) and the linearity of sample means and covariances it follows that 
\begin{align}
  (\#eq:lm-hat-rel)
  \hat \alpha_z &= \hat \alpha_y + \hat \alpha_d \\
  \hat \beta_z &= \hat \beta_y + \hat \beta_d.
\end{align}
That is, the estimated coefficients for $y$, $z$, and $d$ mirror the relationship between the variables themselves,  $z = y + d$.
Consequently, we see that the measurement error $d$ creates absolute errors in the regression coefficients $\hat \alpha$ and $\hat \beta$ equal to
\begin{align}
  (\#eq:lm-hat-err)
  \text{abs. error}(\hat \alpha) &\equiv \hat \alpha_z - \hat \alpha_y = \hat \alpha_d \\
  \text{abs. error}(\hat \beta) &\equiv \hat \beta_z - \hat \beta_y = \hat \beta_d.
\end{align}
<!-- These absolute errors correspond to the _statistical bias_ in the coefficient estimates. -->
Expressed in terms of sample means and covariances, these errors are
\begin{align}
  (\#eq:lm-hat-err-1)
  \text{abs. error}(\hat \alpha) &= \bar d - \hat \beta_d \bar x \\
  \text{abs. error}(\hat \beta)  &= \frac{s(d,x)}{s^2(x)} = r(d,x) \cdot \frac{s(d)}{s(x)}.
\end{align}

@@

Although the absolute error is the same for each species, its practical significance varies.
Recall from Equation \@ref(eq:lm-hat) that the correct value for $\hat \beta$ (i.e., the estimate without measurement error) is
\begin{align}
  \hat \beta &= \frac{s[\log \text{abun}_i, x]}{s^2(x)}.
\end{align}
For species whose log abundance covaries with $x$ much more than the log mean efficiency, the error will be relatively small.
This situation can occur either because the mean efficiency varies relatively little across samples or because its variation is relatively uncorrelated with $x$.
For species whose log abundance covaries with $x$ similar or less than the log mean efficiency, the error will be significant.
A sign error occurs when the species covariance is of _equal sign and smaller in magnitude_ than the covariance of the log mean efficiency.

@@

This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate log fold changes (LFCs) in proportions from MGS count data either with or without bias correction.
